Alone Together: Compositional Reasoning and Inference for Weak Isolation

### 2017-11-08 evaluation.tex
edits

In this section, we present our experience in running \thetool on two
different applications: \emph{Courseware}: a course registration
system described by~\cite{gotsmanpopl16}, and the TPC-C benchmark.

@@

\begin{table}[]
\centering
\begin{tabular}{l|c|c|c|c|c|}
\cline{2-6}
                                 & \multicolumn{1}{l|}{\C{new\_order}} & \multicolumn{1}{l|}{\C{delivery}} & \multicolumn{1}{l|}{\C{payment}} & \multicolumn{1}{l|}{\C{order\_status}} & \multicolumn{1}{l|}{\C{stock\_level}} \\ \hline
\multicolumn{1}{|l|}{MySQL}      & SER                                   & SER                                 & RC                                 & RC                                       & RC                                      \\ \hline
\multicolumn{1}{|l|}{PostgreSQL} & SI                                    & SI                                  & RC                                 & RC                                       & RC                                      \\ \hline
\end{tabular}
\caption{The discovered isolation levels for TPC-C transactions}
\label{tab:tpcc}
\end{table}

>>>>

In this section, we present our experience in running \thetool on two
different applications: \emph{Courseware}: a course registration
system described by~\cite{gotsmanpopl16}, and TPC-C.

@@

\begin{table}[]
\centering
\begin{tabular}{l|c|c|c|c|c|}
\cline{2-6}
                                 & \multicolumn{1}{l|}{\C{new\_order}} & \multicolumn{1}{l|}{\C{delivery}} & \multicolumn{1}{l|}{\C{payment}} & \multicolumn{1}{l|}{\C{order\_status}} & \multicolumn{1}{l|}{\C{stock\_level}} \\ \hline
\multicolumn{1}{|l|}{MySQL}      & SER                                   & SER                                 & RC                                 & RC                                       & RC                                      \\ \hline
\multicolumn{1}{|l|}{PostgreSQL} & SI                                    & SI                                  & RC                                 & RC                                       & RC                                      \\ \hline
\end{tabular}
\caption{The discovered isolation levels for TPC-C transactions}
\label{tab:tpcc}
\vspace*{-10pt}
\end{table}

### 2017-11-08 implementation.tex
edits

At the top-level, \thetool runs a loop that picks an unverified
transaction and progressively strengthens its isolation level until it
passes verification. If the selected data store provides a
serializable isolation level, and if the program is sequentially
correct, then the verification is guaranteed to succeed. Within the
loop, \thetool first computes the various rely relations needed for
verification ($R$, $\R_l$, and $\R_c$). It then traverses the AST of a
transaction, applying the inference rules to construct a state
transformer, checks its stability, and weakens it ($\stabilize{\cdot}$)
if it is not stable. The result of traversing the transaction's AST is
therefore a state transformer ($\F$) that is stable w.r.t $\R_l$, which
is also stabilized against $\R_c$ (using $\stabilize{\cdot}$), and
then checked against the transaction's stated guarantee ($G$). If the
check passes, then the guarantee is verified to check if it preserves
the invariant $I$. The successful result from both checks results
in the transaction being certified correct under the current choice of
its isolation level. Successful verification of all transactions
concludes the top-level execution, returning the inferred isolation
levels as its output.

\thetool uses the Z3 SMT solver as its underlying reasoning engine. Each
implication check described above is first encoded in FOL, applying
the translation described in \S\ref{sec:inference} wherever
necessary.

@@

Clearly, this sequence of events is not atomic, unlike the assumption
made by our formal model because  the implementation admits interference
between the updates of individual records that meet the search
criteria.  Nonetheless, through a series of relatively straightforward
deductions, we can show that PostgreSQL's \C{UPDATE} is in fact
equivalent (in behavior) to a sequential composition of two atomic
operations $c_1;c_2$, where $c_1$ is effectively a \C{SELECT}
operation with the same search criteria as \C{UPDATE}, and $c_2$ is
a slight variation of the original \C{UPDATE} that updates a
record only if a record with the same id is present in the set of records
returned by the \C{SELECT}. This transformation is summarized below:
\begin{smathpar}
\begin{array}{lcl}
\updatee{(\lambda x. ~e_1)}{(\lambda x.~e_2)}
&
\longrightarrow
&
\lete{y}{\selecte{(\lambda x.~e_1})}
     {\updatee{(\lambda x.~e_1 \wedge x.\idf\in\dom(y))}
              {(\lambda x.~e_2})}\\
\end{array}
\end{smathpar}
The intuition behind this translation is the observation that all
interferences possible during the execution of the \C{UPDATE} can be
accommodated between the time the records are selected from the
snapshot, and the time they are actually updated.  \thetool performs this
translation if the selected store is PostgreSQL, allowing it to reason
about \C{UPDATE} operations in a way that is faithful to its semantics
on PostgreSQL. \thetool also admits similar compensatory logic for
certain combinations of isolation levels and operations on MySQL.

\textbf{Set functions} SQL's \C{SELECT} query admits projections of
record fields, and also application of auxiliary functions such as
\C{MAX} and \C{MIN}, e.g., \C{SELECT MAX(ol\_o\_id) FROM
Order\_line WHERE $\ldots$}, etc. We admit such extensions as set functions
in our DSL (e.g., \C{project}, \C{max}, \C{min}), and axiomatize their
behavior. For instance:
\begin{smathpar}
\begin{array}{lcl}
  s_2 \;=\;\C{project}\,s_1\,(\lambda z.~e) & \Leftrightarrow &
  \forall y.~y\in s_2 \Leftrightarrow  \exists(x \in s_1).~y = [x/z]e\\
  x \;=\; \C{max}\,s & \Leftrightarrow & x \in s \conj \forall(y \in
  s).~y\le x\\
\end{array}
\end{smathpar}
There are however certain set functions whose behavior cannot be
completely axiomatized in FOL. These include \C{sum}, \C{count} etc.
For these, we admit imprecise axiomatizations expressed as lemmas
over these functions.

>>>>

At the top-level, \thetool runs a loop that picks an unverified
transaction and progressively strengthens its isolation level until it
passes verification. If the selected data store provides a
serializable isolation level, and if the program is sequentially
correct, then the verification is guaranteed to succeed. Within the
loop, \thetool first computes the various rely relations needed for
verification ($R$, $\R_l$, and $\R_c$). It then traverses the AST of a
transaction, applying the inference rules to construct a state
transformer, checks its stability, and weakens it ($\stabilize{\cdot}$)
if it is not stable. The result of traversing the transaction's AST is
therefore a state transformer ($\F$) that is stable w.r.t $\R_l$, which
is also stabilized against $\R_c$ (using $\stabilize{\cdot}$), and
then checked against the transaction's stated guarantee ($G$). If the
check passes, then the guarantee is verified to check if it preserves
the invariant $I$. The successful result from both checks results
in the transaction being certified correct under the current choice of
its isolation level. Successful verification of all transactions
concludes the top-level execution, returning the inferred isolation
levels as its output.  \thetool uses the Z3 SMT solver as its underlying reasoning engine. Each
implication check described above is first encoded in FOL, applying
the translation described in \S\ref{sec:inference} wherever
necessary.

@@

Clearly, this sequence of events is not atomic, unlike the assumption
made by our formal model because  the implementation admits interference
between the updates of individual records that meet the search
criteria.  Nonetheless, through a series of relatively straightforward
deductions, we can show that PostgreSQL's \C{UPDATE} is in fact
equivalent (in behavior) to a sequential composition of two atomic
operations $c_1;c_2$, where $c_1$ is effectively a \C{SELECT}
operation with the same search criteria as \C{UPDATE}, and $c_2$ is
a slight variation of the original \C{UPDATE} that updates a
record only if a record with the same id is present in the set of records
returned by \C{SELECT}: % This transformation is summarized below:
\begin{smathpar}
\begin{array}{lcl}
\updatee{(\lambda x. ~e_1)}{(\lambda x.~e_2)}
&
\longrightarrow
&
\lete{y}{\selecte{(\lambda x.~e_1})}
     {\updatee{(\lambda x.~e_1 \wedge x.\idf\in\dom(y))}
              {(\lambda x.~e_2})}\\
\end{array}
\end{smathpar}
The intuition behind this translation is the observation that all
interferences possible during the execution of the \C{UPDATE} can be
accommodated between the time the records are selected from the
snapshot, and the time they are actually updated.  \thetool performs this
translation if the selected store is PostgreSQL, allowing it to reason
about \C{UPDATE} operations in a way that is faithful to its semantics
on PostgreSQL. \thetool also admits similar compensatory logic for
certain combinations of isolation levels and operations on MySQL.

\textbf{Set functions} SQL's \C{SELECT} query admits projections of
record fields, and also application of auxiliary functions such as
\C{MAX} and \C{MIN}, e.g., \C{SELECT MAX(ol\_o\_id) FROM
Order\_line WHERE $\ldots$}, etc. We admit such extensions as set functions
in our DSL (e.g., \C{project}, \C{max}, \C{min}), and axiomatize their
behavior. For instance:
\begin{smathpar}
\begin{array}{lcl}
  s_2 \;=\;\C{project}\,s_1\,(\lambda z.~e) & \Leftrightarrow &
  \forall y.~y\in s_2 \Leftrightarrow  \exists(x \in s_1).~y = [x/z]e\\
  x \;=\; \C{max}\,s & \Leftrightarrow & x \in s \conj \forall(y \in
  s).~y\le x\\
\end{array}
\end{smathpar}
There are however certain set functions whose behavior cannot be
completely axiomatized in FOL. These include \C{sum}, \C{count} etc.
For these, we admit imprecise axiomatizations. % expressed as lemmas over these functions.

### 2017-11-08 introduction.tex
edits

The guarantees provided by serializability do not come for free,
however - pessimistic concurrency control methods require databases to
use expensive mechanisms such as two-phase locking that incur overhead
to deal with deadlocks, rollbacks, and
re-execution~\cite{twopl,ullmanbook}.  Similar criticisms apply to
optimistic multi-version concurrency control methods that must deal
with timestamp and version management~\cite{BG81}.  These issues are
further exacerbated when the database is replicated, requiring
additional coordination
mechanisms~\cite{cap,sernotavlbl,bailishat,bernsigmod13}.

>>>>

The guarantees provided by serializability do not come for free,
however - pessimistic concurrency control methods require databases to
use expensive mechanisms such as two-phase locking that incur overhead
to deal with deadlocks, rollbacks, and
re-execution~\cite{twopl,ullmanbook}.  Similar criticisms apply to
optimistic multi-version concurrency control methods that must deal
with timestamp and version management~\cite{BG81}.  These issues are
exacerbated when the database is replicated, requiring
additional coordination
mechanisms~\cite{cap,sernotavlbl,bailishat,bernsigmod13}.

### 2017-11-08 related-work.tex
edits

Recent work described in~\cite{CPA+17} also explores the
use of a state-based interpretation of isolation as we do, and 
like our approach, develops specifications of weak isolation that are not
tied to implementation-specific artifacts.  However, they do not
consider verification (manual or automated) of client programs, and it
is not immediately apparent if their specification formalism is
amenable for use within a verification toolchain.  ~\cite{WB17}
presents a dynamic analysis for weak isolation that attempts to
discover weak isolation anomalies from SQL log files.  Their solution,
while capable of identifying database attacks due to the use of
incorrect (excessively weak) isolation levels, does not consider how
to verify application correctness, infer proper isolation levels, or
formally reason about the relationship between weak-isolation levels
and application invariants.

\paragraph{Reasoning under weak isolation.} ~\cite{feketessi} propose
a theory to characterize non-serializable executions that arise under
{\sc si}. They also propose an algorithm that
allocates either {\sc si} or {\sc ser} isolation levels to
transactions while guaranteeing
serializability. ~\cite{gotsmanpodc16} improve
on Adya's {\sc si} specification and use it to derive a static
analysis that determines the safety of substituting {\sc si} with a
weaker variant called \iso{Parallel Snapshot Isolation}~\cite{psi}.
These efforts focus on establishing the equivalence of executions
between a pair of isolation levels, without taking application
invariants into account.  Bernstein \emph{et al.}~\cite{bern2000}
propose informal semantic conditions to ensure the satisfaction of
application invariants under weaker isolation levels.  All these
techniques are tailor-made for a finite set of well-understood
isolation levels (rooted in~\cite{berenson}).

\paragraph{Reasoning under weak consistency.} There have been several
recent proposals to reason about programs executing under weak
consistency~\cite{bailisvldb, alvarocalm, gotsmanpopl16,redblueatc,
redblueosdi, ecinec}. All of them assume a system model that offers a
choice between a \emph{coordination-free} weak consistency level
(\emph{e.g.}, eventual consistency~\cite{redblueosdi, redblueatc,
ecinec, alvarocalm, bailisvldb}) or causal
consistency~\cite{lbc16,gotsmanpopl16}). All these efforts involve
proving that atomic and fully isolated operations preserve application
invariants when executed under these consistency levels.  In contrast,
our focus in on reasoning in the presence of weakly-isolated
transactions under a strongly consistent store.  ~\cite{gotsmanpopl16}
adapt \iso{Parallel Snapshot Isolation} to a transaction-less setting
by interpreting it as a consistency level that serializes writes to
objects; a dedicated proof rule is developed to help prove prove
program invariants hold under this model. By parameterizing our proof
system over a gamut of weak isolation specifications, we avoid the
need to define a separate proof rule for each new isolation level we
may encounter.

%% \vspace*{-4pt}
%% \paragraph{Reasoning under relaxed memory.} The reasoning mechanisms
%% used to describe and prove properties about weakly-isolated
%% transactions bear some resemblance to those used to formalize relaxed
%% memory behaviour~\cite{battycpp}.  Ridge~\cite{rgtso} generalizes
%% rely-guarantee reasoning to the x86-TSO memory model.  Likewise,
%% Vafeiadis \emph{et al.}~\cite{rsl13} generalize concurrent separation
%% logic (CSL)~\cite{csl} to the C11 relaxed memory model.  Ferreira
%% \emph{et al.}~\cite{ferreira10} propose a parameterized operational
%% semantics for relaxed memory models, but the parameterization is over
%% a relation between relaxed memory programs and related {\sc sc}
%% programs. Demange \emph{et al.}~\cite{DLZ+13} present a \emph{buffered
%%   memory model} for Java that defines an axiomatic definition for the
%% JMM in terms of memory reorderings, and an operational instantiation
%% consistent with the TSO memory model.

\paragraph{Inference.}  ~\cite{Vaf10,Vaf10a}
describe \emph{action inference}, an inference procedure for computing
rely and guarantee relations in the context of RGSep~\cite{VP07}, an
integration of rely-guarantee and separation logic~\cite{Rey02} that
allows one to precisely reason about local and shared state of a
concurrent program. The ideas underlying action inference have been
used to prove memory safety, linearizability, shape invariant
inference, etc.  of fine-grained concurrent data structures.  While
our motivation is similar (automated inference of intermediate
assertions and local invariants), the context of study (transactions
vs. shared-memory concurrency), the objects being analyzed (relational
database tables vs. concurrent data structures), the properties being
verified (integrity constraints over relational tables vs. memory
safety, or linearizability of concurrent data structure operations)
and the analysis technique used to drive inference (state transformers
vs. abstract interpretation) are quite different.

\section{Conclusions}
\label{sec:conclusions}

To improve performance, modern database systems employ techniques that
weaken the strong isolation guarantees provided by serializable
transactions in favor of alternatives that allow a transaction to
witness the effects of other concurrently executing transactions that
happen commit during its execution.  Typically, it is the
responsibility of the database programmer to determine if an available
weak isolation level would violate a transaction's consistency
constraints.  Although this has proven to be a difficult and
error-prone process, there has heretofore been no attempt to formalize
notions of weak isolation with respect to application semantics, or
consider how we might verify the correctness of database programs that
use weakly-isolated transactions.  In this paper, we provide such a
formalization.  We develop a rely-guarantee proof framework cognizant
of weak isolation semantics, and build on this foundation to devise an
inference procedure that facilitates automated verification of
weakly-isolated transactions, and have applied our ideas on
widely-used database systems to justify their utility.  Our solution
enables database applications to leverage the performance advantages
offered by weak isolation, without compromising safety and
correctness.

>>>>

Recent work described in~\cite{CPA+17} also explores the
use of a state-based interpretation of isolation as we do, and 
like our approach, develops specifications of weak isolation that are not
tied to implementation-specific artifacts.  However, they do not
consider verification (manual or automated) of client programs, and it
is not immediately apparent if their specification formalism is
amenable for use within a verification toolchain.  ~\cite{WB17}
presents a dynamic analysis for weak isolation that attempts to
discover weak isolation anomalies from SQL log files.  Their solution,
while capable of identifying database attacks due to the use of
incorrect isolation levels, does not consider how
to verify application correctness, infer proper isolation levels, or
formally reason about the relationship between weak-isolation levels
and application invariants.
\paragraph{Reasoning under weak isolation.} ~\cite{feketessi} propose
a theory to characterize non-serializable executions that arise under
{\sc si}. They also propose an algorithm that
allocates either {\sc si} or {\sc ser} isolation levels to
transactions while guaranteeing
serializability. ~\cite{gotsmanpodc16} improve
on Adya's {\sc si} specification and use it to derive a static
analysis that determines the safety of substituting {\sc si} with a
weaker variant called \iso{Parallel Snapshot Isolation}~\cite{psi}.
These efforts focus on establishing the equivalence of executions
between a pair of isolation levels, without taking application
invariants into account.  Bernstein \emph{et al.}~\cite{bern2000}
propose informal semantic conditions to ensure the satisfaction of
application invariants under weaker isolation levels.  All these
techniques are tailor-made for a finite set of well-understood
isolation levels (rooted in~\cite{berenson}).
\vspace*{-6pt}
\paragraph{Reasoning under weak consistency.} There have been several
recent proposals to reason about programs executing under weak
consistency~\cite{bailisvldb, alvarocalm, gotsmanpopl16,redblueatc,
redblueosdi, ecinec}. All of them assume a system model that offers a
choice between a \emph{coordination-free} weak consistency level
(\emph{e.g.}, eventual consistency~\cite{redblueosdi, redblueatc,
ecinec, alvarocalm, bailisvldb}) or causal
consistency~\cite{lbc16,gotsmanpopl16}). All these efforts involve
proving that atomic and fully isolated operations preserve application
invariants when executed under these consistency levels.  In contrast,
our focus in on reasoning in the presence of weakly-isolated
transactions under a strongly consistent store.  ~\cite{gotsmanpopl16}
adapt \iso{Parallel Snapshot Isolation} to a transaction-less setting
by interpreting it as a consistency level that serializes writes to
objects; a dedicated proof rule is developed to help prove prove
program invariants hold under this model. By parameterizing our proof
system over a gamut of weak isolation specifications, we avoid the
need to define a separate proof rule for each new isolation level we
may encounter.
%% \vspace*{-4pt}
%% \paragraph{Reasoning under relaxed memory.} The reasoning mechanisms
%% used to describe and prove properties about weakly-isolated
%% transactions bear some resemblance to those used to formalize relaxed
%% memory behaviour~\cite{battycpp}.  Ridge~\cite{rgtso} generalizes
%% rely-guarantee reasoning to the x86-TSO memory model.  Likewise,
%% Vafeiadis \emph{et al.}~\cite{rsl13} generalize concurrent separation
%% logic (CSL)~\cite{csl} to the C11 relaxed memory model.  Ferreira
%% \emph{et al.}~\cite{ferreira10} propose a parameterized operational
%% semantics for relaxed memory models, but the parameterization is over
%% a relation between relaxed memory programs and related {\sc sc}
%% programs. Demange \emph{et al.}~\cite{DLZ+13} present a \emph{buffered
%%   memory model} for Java that defines an axiomatic definition for the
%% JMM in terms of memory reorderings, and an operational instantiation
%% consistent with the TSO memory model.
\paragraph{Inference.}  ~\cite{Vaf10,Vaf10a}
describe \emph{action inference}, an inference procedure for computing
rely and guarantee relations in the context of RGSep~\cite{VP07}, an
integration of rely-guarantee and separation logic~\cite{Rey02} that
allows one to precisely reason about local and shared state of a
concurrent program. The ideas underlying action inference have been
used to prove memory safety, linearizability, shape invariant
inference, etc.  of fine-grained concurrent data structures.  While
our motivation is similar (automated inference of intermediate
assertions and local invariants), the context of study (transactions
vs. shared-memory concurrency), the objects being analyzed (relational
database tables vs. concurrent data structures), the properties being
verified (integrity constraints over relational tables vs. memory
safety, or linearizability of concurrent data structure operations)
and the analysis technique used to drive inference (state transformers
vs. abstract interpretation) are quite different.
\vspace*{-7pt}
\section{Conclusions}
\label{sec:conclusions}

To improve performance, modern database systems employ techniques that
weaken the strong isolation guarantees provided by serializable
transactions in favor of alternatives that allow a transaction to
witness the effects of other concurrently executing transactions that
happen commit during its execution.  Typically, it is the
responsibility of the database programmer to determine if an available
weak isolation level would violate a transaction's consistency
constraints.  Although this has proven to be a difficult and
error-prone process, there has heretofore been no attempt to formalize
notions of weak isolation with respect to application semantics, or
consider how we might verify the correctness of database programs that
use weakly-isolated transactions.  In this paper, we provide such a
formalization.  We develop a rely-guarantee proof framework cognizant
of weak isolation semantics, and build on this foundation to devise an
inference procedure that facilitates automated verification of
weakly-isolated transactions, and have applied our ideas on
widely-used database systems to justify their utility.  Our solution
enables database applications to leverage the performance advantages
offered by weak isolation, without compromising correctness.
